---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# measure

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/JamesHWade/measure/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JamesHWade/measure/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/JamesHWade/measure/graph/badge.svg)](https://app.codecov.io/gh/JamesHWade/measure)
<!-- badges: end -->

## Overview

measure extends [tidymodels](https://www.tidymodels.org/) with preprocessing steps for analytical measurement data such as spectroscopy, chromatography, and other instrument-generated signals. It provides a [recipes](https://recipes.tidymodels.org/)-style interface for common spectral preprocessing techniques.

measure helps you:

- **Convert** measurement data from wide or long formats into an internal representation
- **Preprocess** spectra using techniques like smoothing, derivatives, and normalization
- **Transform** data back to wide or long format for modeling or visualization
- **Handle multi-dimensional data** like LC-DAD, EEM fluorescence, and 2D NMR with native nD support
- **Decompose** complex signals using PARAFAC, Tucker, and MCR-ALS methods

## Installation
You can install the development version of measure from [GitHub](https://github.com/):

```r
# install.packages("pak")
pak::pak("JamesHWade/measure")
```

## Usage

The measure workflow follows the familiar recipes pattern: define a recipe, add steps, prep, and bake.

```{r example, message = FALSE}
library(measure)
library(recipes)
library(ggplot2)

# NIR spectroscopy data for predicting meat composition
data(meats_long)
head(meats_long)
```
### Building a preprocessing recipe

```{r recipe}
rec <- recipe(water + fat + protein ~ ., data = meats_long) |>
  # Assign sample ID role (not used as predictor)
  update_role(id, new_role = "id") |>
  # Convert long-format measurements to internal representation
  step_measure_input_long(transmittance, location = vars(channel)) |>
  # Apply Savitzky-Golay smoothing with first derivative
  step_measure_savitzky_golay(window_side = 5, differentiation_order = 1) |>
  # Standard Normal Variate normalization
  step_measure_snv() |>
  # Convert back to wide format for modeling
  step_measure_output_wide(prefix = "nir_")
```

### Preparing and applying the recipe

```{r prep-bake}
# Prep learns any parameters from training data
prepped <- prep(rec)

# Bake applies the transformations
processed <- bake(prepped, new_data = NULL)

# Result is ready for modeling
processed[1:5, 1:8]
```

### Visualizing the preprocessing

```{r visualization, fig.width = 8, fig.height = 4}
# Get data at intermediate step (before output conversion)
rec_for_viz <- recipe(water + fat + protein ~ ., data = meats_long) |>
update_role(id, new_role = "id") |>
  step_measure_input_long(transmittance, location = vars(channel)) |>
  step_measure_savitzky_golay(window_side = 5, differentiation_order = 1) |>
  step_measure_snv()

processed_long <- bake(prep(rec_for_viz), new_data = NULL)

# Extract and plot a few spectra
library(tidyr)
library(dplyr)

plot_data <- processed_long |>
  slice(1:10) |>
  mutate(sample_id = row_number()) |>
  unnest(.measures)

ggplot(plot_data, aes(x = location, y = value, group = sample_id, color = factor(sample_id))) +
  geom_line(alpha = 0.7) +
  labs(
    x = "Channel",
    y = "Preprocessed Signal",
    title = "NIR Spectra After Preprocessing",
    subtitle = "Savitzky-Golay first derivative + SNV normalization",
    color = "Sample"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Available Steps

### Input/Output Steps

| Step | Description |
|------|-------------|
| `step_measure_input_wide()` | Convert wide format (measurements in columns) to internal format |
| `step_measure_input_long()` | Convert long format (measurements in rows) to internal format |
| `step_measure_output_wide()` | Convert back to wide format for modeling |
| `step_measure_output_long()` | Convert back to long format |

### Spectral Math

| Step | Description |
|------|-------------|
| `step_measure_absorbance()` | Convert transmittance to absorbance |
| `step_measure_transmittance()` | Convert absorbance to transmittance |
| `step_measure_log()` | Log transformation with configurable base/offset |
| `step_measure_kubelka_munk()` | Kubelka-Munk transformation for reflectance |
| `step_measure_derivative()` | Simple finite difference derivatives |
| `step_measure_derivative_gap()` | Gap (Norris-Williams) derivatives |

### Filtering & Scatter Correction

| Step | Description |
|------|-------------|
| `step_measure_savitzky_golay()` | Smoothing and/or differentiation |
| `step_measure_snv()` | Standard Normal Variate normalization |
| `step_measure_msc()` | Multiplicative Scatter Correction |
| `step_measure_emsc()` | Extended MSC with wavelength-dependent correction |
| `step_measure_osc()` | Orthogonal Signal Correction |

### Smoothing & Noise Reduction

| Step | Description |
|------|-------------|
| `step_measure_smooth_ma()` | Moving average smoothing |
| `step_measure_smooth_median()` | Median filter (robust to spikes) |
| `step_measure_smooth_gaussian()` | Gaussian kernel smoothing |
| `step_measure_smooth_wavelet()` | Wavelet denoising |
| `step_measure_filter_fourier()` | Fourier low-pass/high-pass filtering |
| `step_measure_despike()` | Spike/outlier detection and removal |

### Sample-wise Normalization

| Step | Description |
|------|-------------|
| `step_measure_normalize_sum()` | Divide by sum (total intensity) |
| `step_measure_normalize_max()` | Divide by maximum value |
| `step_measure_normalize_range()` | Scale to 0-1 range |
| `step_measure_normalize_vector()` | L2/Euclidean normalization |
| `step_measure_normalize_auc()` | Divide by area under curve |
| `step_measure_normalize_peak()` | Normalize by peak region (tunable) |

### Variable-wise Scaling

| Step | Description |
|------|-------------|
| `step_measure_center()` | Mean centering |
| `step_measure_scale_auto()` | Auto-scaling (z-score) |
| `step_measure_scale_pareto()` | Pareto scaling |
| `step_measure_scale_range()` | Range scaling |
| `step_measure_scale_vast()` | VAST scaling |

### Baseline Correction

| Step | Description |
|------|-------------|
| `step_measure_baseline_als()` | Asymmetric least squares |
| `step_measure_baseline_poly()` | Polynomial baseline fitting |
| `step_measure_baseline_rf()` | Rolling ball/LOESS baseline |
| `step_measure_baseline_rolling()` | Rolling ball algorithm |
| `step_measure_baseline_airpls()` | Adaptive Iteratively Reweighted PLS |
| `step_measure_baseline_arpls()` | Asymmetrically Reweighted PLS |
| `step_measure_baseline_snip()` | SNIP (Statistics-sensitive Non-linear Iterative Peak-clipping) |
| `step_measure_baseline_tophat()` | Top-hat morphological filter |
| `step_measure_baseline_morph()` | Iterative morphological correction |
| `step_measure_baseline_minima()` | Local minima interpolation |
| `step_measure_baseline_auto()` | Automatic method selection |
| `step_measure_detrend()` | Polynomial detrending |

### Reference Corrections

| Step | Description |
|------|-------------|
| `step_measure_subtract_blank()` | Blank/background subtraction |
| `step_measure_subtract_reference()` | Reference spectrum subtraction |
| `step_measure_ratio_reference()` | Reference ratio with optional blank |

### Region Operations

| Step | Description |
|------|-------------|
| `step_measure_trim()` | Keep measurements within specified x-range |
| `step_measure_exclude()` | Remove measurements within specified range(s) |
| `step_measure_resample()` | Interpolate to new regular grid |

### Alignment & Registration

| Step | Description |
|------|-------------|
| `step_measure_align_shift()` | Cross-correlation shift alignment |
| `step_measure_align_reference()` | Align to external reference spectrum |
| `step_measure_align_dtw()` | Dynamic Time Warping alignment |
| `step_measure_align_ptw()` | Parametric Time Warping |
| `step_measure_align_cow()` | Correlation Optimized Warping (tunable) |

### Quality Control

| Step | Description |
|------|-------------|
| `step_measure_qc_snr()` | Calculate signal-to-noise ratio |
| `step_measure_qc_saturated()` | Detect saturated measurements |
| `step_measure_qc_outlier()` | Detect outlier samples |
| `step_measure_impute()` | Interpolate missing values |

### Peak Operations

| Step | Description |
|------|-------------|
| `step_measure_peaks_detect()` | Detect peaks using prominence or derivative methods |
| `step_measure_peaks_integrate()` | Calculate peak areas |
| `step_measure_peaks_filter()` | Filter peaks by height, area, or count |
| `step_measure_peaks_deconvolve()` | Deconvolve overlapping peaks |
| `step_measure_peaks_to_table()` | Convert peaks to wide format for modeling |

### SEC/GPC Analysis

| Step | Description |
|------|-------------|
| `step_measure_mw_averages()` | Calculate Mn, Mw, Mz, Mp, and dispersity |
| `step_measure_mw_distribution()` | Generate molecular weight distribution curve |
| `step_measure_mw_fractions()` | Calculate molecular weight fractions |

### Feature Engineering

| Step | Description |
|------|-------------|
| `step_measure_integrals()` | Calculate integrated areas for specified regions |
| `step_measure_ratios()` | Calculate ratios between integrated regions |
| `step_measure_moments()` | Calculate statistical moments from spectra |
| `step_measure_bin()` | Reduce spectrum to fewer points via binning |

### Data Augmentation

| Step | Description |
|------|-------------|
| `step_measure_augment_noise()` | Add random noise for training augmentation |
| `step_measure_augment_shift()` | Random x-axis shifts for shift invariance |
| `step_measure_augment_scale()` | Random intensity scaling |

### Drift & Batch Correction

| Step/Function | Description |
|---------------|-------------|
| `step_measure_drift_qc_loess()` | QC-RLSC drift correction using LOESS |
| `step_measure_drift_linear()` | Linear drift correction |
| `step_measure_drift_spline()` | Spline-based drift correction |
| `step_measure_qc_bracket()` | QC bracketing interpolation |
| `step_measure_batch_reference()` | Reference-based batch correction |
| `measure_detect_drift()` | Detect significant drift in QC samples |

## Analytical Validation Functions

measure provides a comprehensive suite of functions for analytical method validation, designed for compatibility with ICH Q2(R2), ISO 17025, and similar regulatory frameworks.

### Calibration & Quantitation

| Function | Description |
|----------|-------------|
| `measure_calibration_fit()` | Fit weighted calibration curves (linear/quadratic) |
| `measure_calibration_predict()` | Predict concentrations with uncertainty |
| `measure_calibration_verify()` | Continuing calibration verification |
| `measure_lod()` / `measure_loq()` | Detection and quantitation limits |

### Precision & Accuracy

| Function | Description |
|----------|-------------|
| `measure_repeatability()` | Within-run precision |
| `measure_intermediate_precision()` | Between-run precision with variance components |
| `measure_reproducibility()` | Between-lab precision |
| `measure_gage_rr()` | Gage R&R / Measurement System Analysis |
| `measure_accuracy()` | Bias, recovery, and accuracy assessment |
| `measure_linearity()` | Linearity with lack-of-fit testing |
| `measure_carryover()` | Carryover evaluation |

### Method Comparison

| Function | Description |
|----------|-------------|
| `measure_bland_altman()` | Bland-Altman analysis with limits of agreement |
| `measure_deming_regression()` | Deming regression for method comparison |
| `measure_passing_bablok()` | Passing-Bablok non-parametric regression |
| `measure_proficiency_score()` | z-scores, En scores, zeta scores for PT |

### Matrix Effects & Sample Prep QC

| Function/Step | Description |
|---------------|-------------|
| `measure_matrix_effect()` | Quantify ion suppression/enhancement |
| `step_measure_standard_addition()` | Standard addition correction |
| `step_measure_dilution_correct()` | Back-calculate diluted concentrations |
| `step_measure_surrogate_recovery()` | Surrogate/internal standard recovery |

### Uncertainty & Quality Control

| Function | Description |
|----------|-------------|
| `measure_uncertainty_budget()` | ISO GUM uncertainty budgets |
| `measure_uncertainty()` | Combined and expanded uncertainty |
| `measure_control_limits()` | Shewhart, EWMA, or CUSUM limits |
| `measure_control_chart()` | Westgard multi-rule control charts |
| `measure_system_suitability()` | System suitability testing |

### Criteria System

| Function | Description |
|----------|-------------|
| `measure_criteria()` | Define acceptance criteria |
| `measure_assess()` | Evaluate data against criteria |
| `criteria_bioanalytical()` | FDA/EMA bioanalytical presets |
| `criteria_ich_q2()` | ICH Q2 validation presets |

## Learning more

- [Getting Started](https://jameshwade.github.io/measure/articles/measure.html) - A comprehensive introduction to measure
- [Preprocessing Techniques](https://jameshwade.github.io/measure/articles/preprocessing.html) - Deep dive into available preprocessing methods
- [Analytical Validation](https://jameshwade.github.io/measure/articles/validation.html) - Calibration, uncertainty, and method validation

## Datasets

### Included Datasets

The package includes datasets for examples and testing:

| Dataset | Technique | Samples | Description |
|---------|-----------|---------|-------------|
| `meats_long` | NIR | 215 | NIR transmittance spectra of meat samples (from modeldata) |
| `bioreactors_small` | Raman | 210 | Raman spectra from 15 small-scale bioreactors |
| `bioreactors_large` | Raman | 42 | Raman spectra from 3 large-scale bioreactors |
| `hplc_chromatograms` | HPLC-UV | 20 | Simulated HPLC chromatograms with 5 compounds |
| `sec_chromatograms` | SEC/GPC | 10 | Simulated SEC chromatograms (5 standards + 5 polymers) |
| `sec_calibration` | SEC/GPC | 5 | Calibration standards for molecular weight curves |
| `maldi_spectra` | MALDI-TOF | 16 | Simulated mass spectra (4 groups × 4 replicates) |

```r
# Load datasets
data(meats_long)
data(glucose_bioreactors)  # loads bioreactors_small and bioreactors_large
data(hplc_chromatograms)
data(sec_chromatograms)
data(sec_calibration)
data(maldi_spectra)
```

### External Data Sources

For additional test data beyond what's included with measure, these sources provide publicly available analytical measurement data:

**R Packages with Spectral Data:**

| Package | Dataset | Technique | Description |
|---------|---------|-----------|-------------|
| [modeldata](https://modeldata.tidymodels.org/) | `meats` | NIR | Meat composition (wide format version) |
| [prospectr](https://github.com/l-ramirez-lopez/prospectr) | `NIRsoil` | NIR | Soil analysis with 825 samples |
| [ChemoSpec](https://bryanhanson.github.io/ChemoSpec/) | Various | IR, NMR | Multiple spectroscopy datasets |
| [hyperSpec](http://hyperspec.r-forge.r-project.org/) | Various | Raman, IR | Hyperspectral data examples |

```r
# Example: Load NIRsoil from prospectr
# install.packages("prospectr")
data(NIRsoil, package = "prospectr")
```

**Online Repositories:**

- [Mendeley Data](https://data.mendeley.com) - Search "spectroscopy", "chromatography", or "mass spectrometry"
- [Zenodo](https://zenodo.org) - Open science data repository
- [Kaggle Datasets](https://www.kaggle.com/datasets) - Community-contributed datasets
- [NIST Chemistry WebBook](https://webbook.nist.gov/chemistry/) - Reference spectra (IR, MS, UV-Vis)
- [SDBS](https://sdbs.db.aist.go.jp/) - Spectral Database for Organic Compounds (NMR, IR, MS)

**Domain-Specific Databases:**

| Database | Data Type | URL |
|----------|-----------|-----|
| MassBank | Mass spectra | https://massbank.eu/MassBank/ |
| HMDB | NMR, MS metabolomics | https://hmdb.ca/ |
| NMRShiftDB | NMR spectra | https://nmrshiftdb.nmr.uni-koeln.de/ |
| Crystallography Open Database | XRD patterns | https://www.crystallography.net/cod/ |

## Related packages

measure builds on the [tidymodels](https://www.tidymodels.org/) ecosystem:

- [recipes](https://recipes.tidymodels.org/) - The foundation for preprocessing pipelines
- [parsnip](https://parsnip.tidymodels.org/) - Unified modeling interface
- [workflows](https://workflows.tidymodels.org/) - Bundle preprocessing and modeling
- [tune](https://tune.tidymodels.org/) - Hyperparameter tuning (works with measure's tunable steps!)

For spectral analysis in R, you might also find these packages useful:

- [prospectr](https://github.com/l-ramirez-lopez/prospectr) - Spectral preprocessing functions
- [ChemoSpec](https://bryanhanson.github.io/ChemoSpec/) - Exploratory chemometrics
- [mdatools](https://mdatools.com/) - Multivariate data analysis

## Contributing

This package is under active development. Contributions are welcome! Please see the [contributing guidelines](https://jameshwade.github.io/measure/CONTRIBUTING.html).

## Code of Conduct

Please note that the measure project is released with a [Contributor Code of Conduct](https://jameshwade.github.io/measure/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
