---
title: "Preprocessing Techniques"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing Techniques}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE
)
```

```{r setup, message = FALSE}
library(measure)
library(recipes)
library(dplyr)
library(tidyr)
library(ggplot2)
library(modeldata)

# Helper function to process and plot spectra
plot_spectra <- function(data, title, subtitle = NULL) {
  ggplot(data, aes(x = location, y = value, group = sample_id, color = factor(sample_id))) +
    geom_line(alpha = 0.7, linewidth = 0.5) +
    labs(x = "Wavelength", y = "Signal", title = title, subtitle = subtitle) +
    theme_minimal() +
    theme(legend.position = "none")
}

# Prepare sample data
data(meats)
wavelengths <- seq(850, 1050, length.out = 100)

# Get spectra in internal format for demonstrations
get_internal <- function(rec) {
  bake(prep(rec), new_data = NULL) |>
    slice(1:15) |>
    mutate(sample_id = row_number()) |>
    unnest(.measures)
}
```

## Introduction

Spectral preprocessing is essential for building accurate chemometric models. Raw spectra often contain unwanted variation from physical effects (scatter, baseline drift) that obscure the chemical information we're trying to model. This vignette covers each preprocessing technique available in measure and when to use them.

## Why preprocess spectra?

Before diving into specific techniques, let's understand what we're dealing with. Here are raw NIR spectra from the meats dataset:

```{r raw-spectra}
rec_raw <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths)

raw_data <- get_internal(rec_raw)

plot_spectra(raw_data, "Raw NIR Spectra", "Note the vertical offset differences between samples")
```

Notice how spectra are shifted vertically relative to each other? This offset isn't due to chemical differences - it's caused by physical factors like particle size, path length, and light scatter. Our preprocessing goal is to remove these unwanted effects while preserving the chemical information.

## Savitzky-Golay Filtering

### What it does

The Savitzky-Golay filter performs polynomial smoothing and can compute derivatives. It fits a polynomial to a sliding window of points, using the polynomial's value (or derivative) at the center point as the output.

### When to use it

- **Smoothing (order = 0)**: Reduce random noise while preserving peak shapes
- **First derivative (order = 1)**: Remove constant baseline offsets, enhance peak differences
- **Second derivative (order = 2)**: Remove linear baseline trends, further enhance peak resolution

### Parameters

- `window_side`: Number of points on each side of the center point (total window = 2 * window_side + 1)
- `differentiation_order`: 0 for smoothing, 1 for first derivative, 2 for second derivative
- `degree`: Polynomial degree (defaults to differentiation_order + 1)

### Examples

```{r sg-examples}
# Just smoothing
rec_smooth <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
  step_measure_savitzky_golay(window_side = 7, differentiation_order = 0)

# First derivative
rec_d1 <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
  step_measure_savitzky_golay(window_side = 5, differentiation_order = 1)

# Second derivative
rec_d2 <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
  step_measure_savitzky_golay(window_side = 7, differentiation_order = 2)
```

```{r sg-plots, fig.height = 8}
library(patchwork)

p1 <- plot_spectra(raw_data, "Raw")
p2 <- plot_spectra(get_internal(rec_smooth), "Smoothed (window = 15)")
p3 <- plot_spectra(get_internal(rec_d1), "1st Derivative", "Baseline offset removed")
p4 <- plot_spectra(get_internal(rec_d2), "2nd Derivative", "Linear baseline removed")

(p1 + p2) / (p3 + p4)
```

### Choosing window size

The window size is a bias-variance trade-off:
- **Smaller window**: Less smoothing, preserves sharp features, more noise
- **Larger window**: More smoothing, may blur sharp peaks, less noise

A good starting point is a window that spans the narrowest feature you want to preserve.

```{r window-comparison}
windows <- c(3, 7, 15)

window_data <- lapply(windows, function(w) {
  rec <- recipe(water ~ ., data = meats) |>
    step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
    step_measure_savitzky_golay(window_side = w, differentiation_order = 1)

  get_internal(rec) |>
    filter(sample_id == 1) |>
    mutate(window = paste0("window_side = ", w))
}) |>
  bind_rows()

ggplot(window_data, aes(x = location, y = value, color = window)) +
  geom_line() +
  labs(
    x = "Wavelength",
    y = "Signal",
    title = "Effect of Window Size on First Derivative",
    color = NULL
  ) +
  theme_minimal()
```

### Tuning with dials

The Savitzky-Golay step is tunable! This means you can use `tune()` to find optimal parameters:

```{r tunable, eval = FALSE}
library(tune)
library(workflows)

rec_tunable <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_")) |>
  step_measure_savitzky_golay(
    window_side = tune(),
    differentiation_order = tune()
  ) |>
  step_measure_output_wide()

# The tunable parameters are:
tunable(rec_tunable)
```

## Standard Normal Variate (SNV)

### What it does

SNV normalizes each spectrum independently by centering and scaling:

$$SNV(x) = \frac{x - \bar{x}}{s_x}$$

where $\bar{x}$ is the spectrum's mean and $s_x$ is its standard deviation.

### When to use it

- Remove multiplicative scatter effects
- Correct for path length variations
- Normalize spectra to similar magnitude

SNV is particularly effective for diffuse reflectance spectra where particle size causes scatter variations.

### Example

```{r snv-example}
rec_snv <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
  step_measure_snv()

snv_data <- get_internal(rec_snv)
plot_spectra(snv_data, "After SNV Normalization", "Each spectrum has mean = 0 and sd = 1")
```

### Combining with derivatives

SNV is often combined with Savitzky-Golay derivatives. The order matters:

```{r snv-combination}
# Derivative then SNV (more common)
rec_d1_snv <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
  step_measure_savitzky_golay(window_side = 5, differentiation_order = 1) |>
  step_measure_snv()

plot_spectra(get_internal(rec_d1_snv), "1st Derivative + SNV",
             "Combined baseline removal and scatter correction")
```

## Multiplicative Scatter Correction (MSC)

### What it does

MSC aligns each spectrum to a reference spectrum (typically the mean of all training spectra) by correcting for additive and multiplicative effects:

1. Fit each spectrum $x_i$ to the reference $x_r$: $x_i = m_i \cdot x_r + a_i$
2. Correct: $MSC(x_i) = \frac{x_i - a_i}{m_i}$

### When to use it

- Similar applications to SNV
- When you have a good reference spectrum
- Often slightly better than SNV for scatter correction

### How it differs from SNV

- **SNV**: Each spectrum normalized independently (no reference needed)
- **MSC**: All spectra aligned to a common reference (learns reference during prep)

This means MSC is a *trained* step - it learns the reference spectrum from training data and applies the same reference to new data.

### Example

```{r msc-example}
rec_msc <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_"), location_values = wavelengths) |>
  step_measure_msc()

msc_data <- get_internal(rec_msc)
plot_spectra(msc_data, "After MSC", "Spectra aligned to mean reference")
```

### Comparing SNV and MSC

```{r snv-vs-msc, fig.height = 6}
p_snv <- plot_spectra(get_internal(rec_snv), "SNV")
p_msc <- plot_spectra(get_internal(rec_msc), "MSC")

p_snv / p_msc
```

Both methods produce similar results for this dataset. In practice, try both and compare model performance.

## Preprocessing pipelines

### Common combinations

Here are some commonly used preprocessing pipelines:

```{r pipelines}
# Pipeline 1: Basic scatter correction
pipe1 <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_")) |>
  step_measure_snv() |>
  step_measure_output_wide()

# Pipeline 2: Derivative + normalization
pipe2 <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_")) |>
  step_measure_savitzky_golay(window_side = 5, differentiation_order = 1) |>
  step_measure_snv() |>
  step_measure_output_wide()

# Pipeline 3: Second derivative (often enough on its own)
pipe3 <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_")) |>
  step_measure_savitzky_golay(window_side = 7, differentiation_order = 2) |>
  step_measure_output_wide()

# Pipeline 4: MSC + smoothing
pipe4 <- recipe(water ~ ., data = meats) |>
  step_measure_input_wide(starts_with("x_")) |>
  step_measure_msc() |>
  step_measure_savitzky_golay(window_side = 5, differentiation_order = 0) |>
  step_measure_output_wide()
```

### Order of operations

The order of preprocessing steps matters. General guidelines:

1. **Derivatives before normalization**: Apply Savitzky-Golay derivatives first, then SNV/MSC
2. **Smoothing after scatter correction**: If using smoothing (not derivatives), apply after MSC/SNV
3. **Keep it simple**: Often, a single well-chosen step outperforms complex pipelines

## Summary table

| Step | Effect | Use when |
|------|--------|----------|
| `step_measure_savitzky_golay(order=0)` | Smoothing | High-frequency noise |
| `step_measure_savitzky_golay(order=1)` | 1st derivative | Baseline offsets |
| `step_measure_savitzky_golay(order=2)` | 2nd derivative | Linear baselines |
| `step_measure_snv()` | Row normalization | Scatter, path length |
| `step_measure_msc()` | Align to reference | Scatter (supervised) |

## Tips for choosing preprocessing

1. **Start simple**: Try SNV or first derivative alone before complex pipelines
2. **Visualize**: Always plot preprocessed spectra to check for artifacts
3. **Validate**: Use cross-validation to compare preprocessing strategies
4. **Domain knowledge**: Consider the physics of your measurement system
5. **Tune**: Use `tune()` to optimize Savitzky-Golay parameters

## References

- Savitzky, A., and Golay, M. J. E. (1964). Smoothing and Differentiation of Data by Simplified Least Squares Procedures. *Analytical Chemistry*, 36(8), 1627-1639.
- Barnes, R. J., Dhanoa, M. S., and Lister, S. J. (1989). Standard Normal Variate Transformation and De-Trending of Near-Infrared Diffuse Reflectance Spectra. *Applied Spectroscopy*, 43(5), 772-777.
- Geladi, P., MacDougall, D., and Martens, H. (1985). Linearization and Scatter-Correction for Near-Infrared Reflectance Spectra of Meat. *Applied Spectroscopy*, 39(3), 491-500.
