---
title: "What is a recipe?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What is a recipe?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidymodels)
library(measure)
```


```{r}
data("credit_data")

set.seed(55)
train_test_split <- initial_split(credit_data)

credit_train <- training(train_test_split)
credit_test <- testing(train_test_split)
```

### Creating a Recipe

We specify a recipe providing formula and data arguments. Check out [*Tidy Modeling with R*](https://www.tmwr.org/base-r.html) to learn more about specifying formulas in R.

```{r}
rec_obj <- recipe(Status ~ ., data = credit_train)
```

The `recipe` funtion returns a recipe object. The formula argument determines the roles of each variables. `Status` is assigned the role of `outcome`, and the 13  other variables are assigned to role of `predictor`.

```{r}
rec_obj
```

Diving a bit deeper, the recipe object is a list with 7 elements. Within these elements, we can see more details are saved about our variables. This includes the `type` and `source` stored in `rec_obj$var_info`.

```{r}
cat(names(rec_obj), sep = "\n")
rec_obj$var_info
```

### Adding a Step

The recipe does not yet contain any steps.

```{r}
rec_obj$steps

rec_obj_add_step <- rec_obj |> 
  step_impute_knn(all_predictors())

rec_obj_add_step$steps
```

